---
title: "Final Project Story"
author: "Arjun"
date: "2023-12-09"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

**Introduction**
<br>
Have you ever wondered why students are the biggest group of consumers of caffeine? It's more than just a habit, especially during exams where the environment of the academic stress and workload is more intense than ever. We've all been there, the sleepless nights, the pressure, the need for an extra kick to power through. That's why we've chosen students as our focus. They are the biggest consumers of caffeine. For this project our intended audience encompasses a big range of students. Those interested in utilizing the dataset are from a range of staff. professors, and school administrators seeking to develop a deeper understanding of the impact of caffeine on students' focus in both school and the classroom and to get better educated on the entire project. The datasets we cover are on the general effects of caffeine on students. The information was gathered through research and surveys conducted by students across various school and university campuses. From the research we found it included the side effects of caffeine on students, the different types of caffeine, effects on sleep and sleep schedule, etc.  Especially in an area like Seattle, coffee is everywhere you turn, and the cafes are endless. Looking over all of this, we wanted to be able to dive deeper on the effects of caffeine and the side effects on students specifically and how to help maintain it. 
<br>
 	With students constantly navigating academic stress and heavy workloads, many turn to caffeine as a coping mechanism. We chose this specific topic as we felt like it was a very relatable topic for many people, especially with us being students who are also going through this same issue of caffeine. As we have seen and interacted with many people and communities on UW campus, we can empathize with the role caffeine plays in their lives, and also the amount of data from many other schools with the same topic. 
Our target audience is mostly focused on students and the datasets we are working with focus on understanding the general impact of caffeine on students. Data collection is made through research and surveys conducted across various school and university campuses. 
Despite widespread awareness of these side effects, individuals continue to willingly subject their bodies to caffeine. Our objective is to go into detail into the side effects of caffeine and ways to help manage and maintain that as a student. 
Ultimately, through this project and research, we aim to provide a deeper insight into the consequences of excessive caffeine consumption among students. We hope to offer recommendations for managing or limiting caffeine usage in a healthy manner, recognizing the importance of balancing academic demands as well as with overall well-being.
<br>
**Explain your Data**
  <br>
	We decided to use these datasets, as it was a confirmed study on the effects of caffeine across over 100 students. There was various data presented in the study, which made it easier to find the topics/correct data we needed to answer our question. The various data included data on the backgrounds of each student, the amount of caffeine consumed, the time of day it was consumed, and the resulting effects of caffeine. With a wide range of data, we were able to evaluate the data that was relevant to our research, and merge it into one dataset to show our trends.
<br>
Our datasets originated from a cross-sectional study that was conducted in the UAE from December 2019 to March 2020. The study observed 500 students from different universities in the UAE. The data was studied and organized by students at the Al Ain University, under the guidance of Dr. Azza Ramadan. 
The data is freely available for everyone to look at, and sampling is fine as long as there is credit and it is for non-commercial purposes. In order to access the complete, raw, dataset, one needs to request permission from the researchers by directly contacting them.
<br>
	The datasets are about every aspect of students’ caffeine consumption and the effects of their consumption. There are multiple datasets just regarding the background/demographic of the various test participants. There were also various datasets that summarized the nature of the experiments, such as how much caffeine was taken, when, and why, as well as the resulting side effects.
<br>	
	The datasets are about the consumption of caffeine amongst University students. The population of this experiment are University students who drink caffeine. The data only encompasses 500 cases of such individuals, which is a very small sample in the overall picture. Though the 500 individuals are all from different backgrounds and physical stature, it is important to note that the data may still be skewed or biased due to the lack of the region– all 500 students were from different schools that were all in the UAE. Therefore the data may have more validity in reference to the region compared to globally. Also, the proportionately small data sample compared to the overall population may lead to misrepresentation. It’s crucial to question whether it is fair for the research to generalize the effects of caffeine on university students based on the small sample size they studied.
<br>	
	Overall, the data quality wasn’t bad in terms of just the individual presentations. However, each data set was formatted differently, and the basis for each data was different as well. In some data charts, each row would be a type of effect, with the numbers as students, while in others, each row is a specific case of each student. This made it difficult to sort out and determine which data was actually relevant to answering our main research question. It was difficult to find what the purpose of each individual data set was, as well as differentiate between the sets. This added confusion and inconsistency in data did cause some issues in reshaping and organizing the data to make it more relevant to us.
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(rvest)
```

```{r}
library(readxl)
caffeine_dataset_1 <- read_excel("../Documents/caffeine dataset 1.xlsx", 
    sheet = "Sheet2")
```

```{r}
library(readxl)
caffeine_dataset_2 <- read_excel("../Documents/caffeine dataset 2.xlsx", 
    sheet = "Sheet2")
```

```{r}
caffeine_dataset_2_longer <- caffeine_dataset_2 %>%
  pivot_longer(c(`8 to 12`, `13 to 17`, `Over 18`, `Under 8`),
               names_to="Age",
               values_to="Count_Caffeine")
```

```{r}
caffeine_dataset_1 %>%
  ggplot(aes(x = Age, y = `Caffeine Consumption`)) +
  geom_bar(stat = "identity", fill = "blue", width = 0.5) +
  labs(title = "Caffeine Consumption by Age",
       x = "Age",
       y = "Caffeine Consumption") +
  annotate("text", x = Inf, y = Inf, hjust = 1, vjust = 1, 
           label = paste("Total Participants: 467 "),
           size = 4, color = "darkblue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))
```

```{r}
ggplot(caffeine_dataset_2_longer, aes(x = Parameter, y = Count_Caffeine, fill = Age)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7) , color = "black", size = 0.7) +
  labs(title = "Caffeine Consumption by Parameter and Age Group",
       x = "Parameter",
       y = "Count of Caffeine") +
  scale_fill_manual(values = c('13 to 17' = 'blue', 'Over 18' = 'red')) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
<br>
Based on data 1, we can see that age 20-23 tends to consume most caffeine (263, However Based on graph 2, we can see that headaches(13-17) and drowsiness(13-17) are the most common side effects of caffeine consumption, which could potentially means that age 13-17 are prone to coffee side effects.

```{r}
library(readxl)
caffeine_dataset_5 <- read_excel("../Documents/caffeine dataset 5.xlsx")
caffeine_dataset_5
caffeine_dataset_5 %>% 
  ggplot(aes(Reason, Percent))+
  geom_col(color = "navy", fill="lavender")
```
<br>
From the data we collected, the top reasons behind caffeine consumption in University students was sleepiness and studying, in which over half of the students who participated in the data agreed to. Reasons such as driving, headaches, and homework were also common, though in this case, the numbers seemed a bit more even amongst the topics.

```{r}
library(readxl)
caffeine_dataset_3 <- read_excel("../Documents/caffeine dataset 3.xlsx") 
caffeine_dataset_3_clean <- caffeine_dataset_3 %>% 
  select(Frequency, Amount, Sources, `N%...10`)
```

```{r}
caffeine_dataset_3_clean %>%
  ggplot(aes(Frequency, Amount))+
  geom_col(fill = "lightblue", col ="pink")+
  labs(title="Caffeine Consumption Frequency vs. Amount of People")
```
<br>
The data displays the amount of times people consume caffeine daily. The graph shows that most people consume caffeine 2-3 times a day at 180 people and the next highest in consumption is once daily at 116 people. As the frequency of daily caffeine consumption increases, for the most part the amount of people decreases. The sweet spot for consumption seems to be between 1-3 times a day.
```{r}
caffeine_dataset_3_clean %>% 
  drop_na(Sources) %>% 
  ggplot(aes(Sources, `N%...10`))+
  geom_col(fill = "brown", col ="pink")+
  labs(title="Caffeine Consumption Sources", y="Frequency")
```
<br>
The graph displays the different sources of caffeine and their popularity among the test subjects. Note our scale has an error so 77 should be at the bottom and not the top. Therefore coffee is the actual most popular source of caffeine at 316 people drinking it to get their caffeine intake.

```{r}
library(readxl)
dataset_coffee_4 <- read_excel("../Downloads/dataset_4.xlsx")
numerical_value <- c(23.93,60.39,12.93,2.76,81.02,14.72,2.89,1.38,44.84,37.83, 12.24,5.09)  
dataset_coffee_4 <- dataset_coffee_4 %>% 
  mutate(Percentage = numerical_value) %>% 
  mutate(percent_50 = Percentage > 50)
```

```{r}
combined <- inner_join(caffeine_dataset_1, dataset_coffee_4,
                       by = c("Total"="Total"))
combined_total <- combined %>% 
  mutate(Total = c(727, integer(32)))
```

```{r}
combined_total %>% 
  group_by(Age) %>% 
  ggplot(aes(Unit,Percentage, fill = factor(Variable)))+
  geom_col(position="dodge")+
  labs(title="Amount of Coffee Consumed vs. Percentage of Participants", x = "Units of Coffee")
```
<br>
The graph of our final merged dataset displays the percentage of participants that consume x amounts of differing types of cofffee. Based on this dataset, 80% of people don't consume decaf coffee and around 45% for concentrated coffee. The majority of people opt for 1-2 units of regular coffee.
<br>
Discuss Your Methods
	For our data processing methods, we first determined which of the data from the original source was actually relevant to our topic question. To determine this, we reflected back on the guiding questions we came up with, and determined what the main variables for each question were. For example, the main variables for the question regarding how caffeine consumption differs by age is based on the variables age and amount of caffeine consumed. We would then look for the data that encompasses these two variables, then determine which type of plot would best show the trend. For our overall data, the variables we used included total (number of participants), age, and caffeine consumption (in units). We included these variables, as we thought this data was the most relevant to depicting the trend that answers our main topic of caffeine consumption in university students.
	<br>
	Before we processed our data, we needed to reformat all of the information to be easier to work with in R Studio. The original data was all in a data table, which was impossible to extract, so our group had to reorganize the data into a Excel spreadsheet in order to be able to upload it onto R Studio as a CSV file. From then on, we decided which columns would be the easiest key for our two datasets to merge by, and merged the data into the overall data we have now. 
	<br>
	We had to make many technical decisions to manipulate the data in a way it would show a proper graph when we coded it on Studio R. As mentioned previously, we had to reorganize all of the data onto an Excel spreadsheet from the given data tables. As for the data itself, certain columns weren’t showing properly when run, which meant we had to delete the issue columns and manually code the data into the dataframe as new vector columns. For certain variables such as age, the variable was not consistent for the whole demographic. For example, certain variables would contain the data for male students within a certain age group, but leave out the female students. To combat this, depending on the data size and condition, we either combined multiple groupings of data into one, or marked them N/A, and excluded the data from our final dataframe. Any missings that existed in individual observations were also removed from our final evaluation of our findings.
	<br>
	We didn’t have statistical methods that were too complex or out of our way, as most of the complexity in our project lay in the cleaning, formatting, and manipulation of our data in a way that would be fair and presentable with as least bias as possible. It was difficult to spot bias in a large dataframe with little context, and we spent a majority of our energy on identifying and finding ways we could remove or neutralize the data. In the end, I would say our methods were logically thought out, and we made what we believe were the correct technical decisions in order to present what we believed were fair findings.

